---
title: "Data 583: Project Proposal"
author: "Ricky Heinrich & Vimaljeet Singh"
date: "`r Sys.Date()`"
output: html_document
---

The data set is taken from UCI machine learning repository and related to the two-year historical log corresponding to years 2011 and 2012 from Capital Bikeshare system, Washington D.C., USA which is publicly available at http://capitalbikeshare.com/system-data. The data was aggregated on two hourly and daily basis. Then the corresponding weather and seasonal information was added on to this data. Weather information are extracted from http://www.freemeteo.com. 

```{r}
day = read.csv("day.csv", header = TRUE)
str(day)
```

Underlying scientific processes that affected the data would be the occurrence of natural events like Hurricane Sandy on Oct 30, 2012. Other factors that might affect the numbers would be hourly or daily environmental and seasonal settings.

















```{r}
df <- data.frame(
  Name = c("John", "Jane", "Bob"),
  Age = c(30, 25, 40),
  Gender = c("Male", "Female", "Male")
)

# Create table
knitr::kable(df)
```

